Statically Typed Web Programming using Scala

### 2012-12-13 papers/ecoop2013/repairing.tex
Implementation section draft

\title{Statically typed Web programming}

\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered} \and Olivier \noun{Barais}\textsuperscript{\textasteriskcentered} \and
Tiark \noun{Rompf}\textsuperscript{\textdagger} \and Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France: \texttt{\{first\}.\{last\}@irisa.fr}\\
\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{\{first\}.\{last\}@epfl.ch}}

@@

\begin{abstract}
Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and
asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}). We investigate
further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging. We define
safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to
efficient JavaScript code using native APIs. We use both type-level information and staging to introduce high-level
abstractions with few or no overhead on the generated JavaScript code.
\end{abstract}

@@

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
writing distributed code requires its parts to be consistent together and leads to duplication unless one uses the
same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing,
no module system, verbose syntax, \etc). Some other in-browser execution environments give the opportunity to write
the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe
Flash~\cite{Curtis00_Flash} and Microsoft
Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not
be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
structured in URLs that users can bookmark or share.

An increasing number of initiatives attempt to allow developers to write the client-side code in a language different
than JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). These languages are
usually more suitable to write large applications either because they provide more constructs to build abstractions
(such as object orientation), or because they are statically typed, or because they add some useful concepts
that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile
for both server and client sides usually allows developers to share some parts of code between server and client
sides for free. However this shared code is restricted to use exclusively language constructs: concepts provided by
external APIs can't be shared between server and client sides because the bindings with these environments are
different on the server and client sides.

In this paper we investigate further a path already introduced by Kossakowski \etal{}: defining JavaScript as a
compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}. This approach is based on Lightweight Modular
Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate
representation that can be further processed to perform domain specific optimizations and then to generate a
JavaScript program (for the client-side) and a Scala program (for the server-side). This approach has two main
advantages: (1) the embedding reduces the effort needed to define the language since we can reuse the Scala
infrastructure and tooling, and (2) staging gives more knowledge to the compiler about how a given abstraction
should be efficiently translated into the target environment. In other words any library-level abstraction can have
the efficiency of a language-level abstraction.

The previous work showed how to write client-side code in Scala using the JavaScript embedded DSL and how to share
code between server-side and client-side. The authors also showed that the code was safer and more convenient to
write thanks to Scala's static typing and class system. Finally, they showed how to \emph{un-invert} the control of
callback-based asynchronous APIs by using continuations but they did not address other concerns of RIAs development
such as user interface events handling, DOM creation and manipulation, and application’s state management.

This paper continues their work and presents safe and expressive DSL constructs for RIAs development. We increase the
expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs. We
make asynchronous programming simpler by raising asynchronous computations to first-class citizen. We leverage the
Scala’s type system to bring type-directed language features such as \emph{adhoc} polymorphism and type coercion to
the client-side.

@@

The next section introduces a simple application example showing challenges of RIA development, section
\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution and
compares it to related work and section \ref{discussion} concludes.

@@

\section{Background and motivating example}
\label{problem}

To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind
mapping” application and show how the client-side code would be typically written in JavaScript.

\subsection{Example description}

In our application, a mind map is represented as a graph where ideas or concepts are represented by vertices, and the
relation between two ideas is represented by an edge. Users can visualize a mind map, zoom in and out and move it.
They can also add ideas and link them together. The figure \ref{mindmap} shows the visualization of a simple map
consisting of three ideas linked together, using our application.

@@

\subsubsection{Events handling}

@@

We think this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
the listened event but developers have no way to check that the fields they’re using are indeed defined on the event
they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the
\code{mousewheel} event.

@@

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or
Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}}
aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure
and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
client-side and server-side, however none has a practical expression language (dynamic content can only be provided
as a map of key-value pairs and no operation can be done on values) or is statically typed.

@@

\subsubsection{Lack of static checks}

Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
and robust code bases and to perform refactorings. For instance, in the case of RIAs, a mispell in an event name may
break the program behavior without getting a sensible error message.

@@

Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
bothers modularization. Furthermore, native APIs are too verbose, making the code hard to read.

@@

\section{DSL design and implementation}
\label{solution}

@@

@@

The remaining of this section describe the design and the implementation of DSLs for Web programming using LMS.

\subsection{Events handling}

We want to bring safety to the events handling API so that developers can’t mispell an event name and can’t attempt
to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
type of the event passed to the handler varies with the event name. For instance in the listing \ref{event-js}, the
handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
\code{MouseWheelEvent} have a property \code{wheelDeltaY} but that’s not the case of an event data of type
\code{KeyboardEvent}, for example.

@@

However, implementing this solution requires a high effort because we have to define as many methods as there are
events. We want to provide a single polymorphic method similar to the native API. To achieve that we need to encode
at the type-level the dependency relation between an event name and its corresponding event type. This dependency can
naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:

@@

\subsection{Asynchronous programming}

@@

We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods
turning them into first-class citizen. The DSL is monadic so we can solve the callback-hell problem thanks to the
Scala \code{for} notation.

@@

An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
\begin{itemize}
\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} eventually transforms the value when available~;
\end{itemize}

Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class
citizen: functions can take as parameters and return asynchronous values.

@@

Listing \ref{async-parallel-1} and \ref{async-parallel-2} define two asynchronous computations running in parallel
and the generated JavaScript code (we renamed some identifiers for the sake of readability). The generated
\code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the
asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a
\code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when
received.

% \begin{lstlisting}[caption=Concrete syntax of the asynchronous programming DSL]
% trait FutureOps { this: Base =>
%   implicit def FutureOpsCls[A](f: Rep[Future[A]]): FutureOpsCls[A]
%   type FutureOpsCls[+A] <: FutureOpsBase[A]
%   trait FutureOpsBase[+A] {
%     def map[B](f: Rep[A] => Rep[B]): Rep[Future[B]]
%     def flatMap[B](f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]
%     def foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]
%     def withFilter(f: Rep[A] => Rep[Boolean]): Rep[Future[A]]
% The \code{FutureOps} module depends on the \code{Base} module defined by LMS. The \code{Base} module essentially defines the abstract type constructor \code{Rep[_]}. Values of type \code{Rep[Future[A]]} have no methods because the \code{Rep[_]} type is abstract and unbound, so we use an implicit conversion from \code{Rep[Future[A]]} to \code{FutureOpsCls[A]} to add methods to \code{Rep[Future[A]]} values. The type \code{FutureOpsCls[A]} is bound by the \code{FutureOpsBase[A]} trait that defines the signature of the methods added to \code{Rep[Future[A]]} values. Everything in the \code{FutureOps} module is abstract, these definitions only provide the vocabulary of the DSL.
% The listing \ref{futureopsexp} shows the definition of the \code{FutureOpsExp} module that implements the \code{FutureOps} module and defines how statement nodes are built from the DSL constructs.
% \begin{figure}
% \begin{lstlisting}[label=futureopsexp,caption=Definition of the intermediate representation of the asynchronous programming DSL]
% trait FutureOpsExp extends FutureOps { this: EffectExp =>
%   implicit class FutureOpsCls[+A](a: Exp[Future[A]])
%       extends FutureOpsBase[A] {
%     def map[B](f: Exp[A] => Exp[B]) = {
%       val x = fresh[A]
%       val b = reifyEffects(f(x))
%       reflectEffect(FutureMap(a, x, b), summarizeEffects(b).star)
%     // ... same for other methods (flatMap, foreach and withFilter)
%   case class FutureMap[A, B](a: Exp[Future[A]], x: Sym[A], b: Block[B])
%       extends Def[Future[B]]
%   // ... same for other operations
% \end{figure}
% The \code{FutureOpsExp} module depends on the \code{EffectExp} module that is a module refining the \code{Base} module, defining the type \code{Rep[A]} as \code{Exp[A]}, which is a base class defining intermediate representations as a graph of statements. The \code{EffectExp} module also provides tools to handle effectful statements. The implicit class \code{FutureOpsCls} implements both the abstract type and the implicit conversion of the same name defined in the \code{FutureOps} trait. Each method of the \code{FutureOpsCls} class is reified to a statement node, for instance the \code{map} method returns a \code{FutureMap} node containing a reference to the node on which the \code{map} operation is applied and the code block that was passed as a parameter to the \code{map} operation. We propagate the information about the effects performed by the statements in the code block by calling \code{reflectEffect} on our statement node.
% The listing \ref{jsgenfutureops} shows the code of the JavaScript code generator for the DSL. The code generation process basically consists in traversing the graph of statement nodes and in emitting the code corresponding to each node. To implement a code generator for a given DSL we just need to say how to generate code for this DSL statement nodes.
% \begin{figure}
%  \begin{lstlisting}[label=jsgenfutureops,caption=JavaScript code generator for the asynchronous programming DSL]
% trait JSGenFutureOps extends JSNestedCodegen {
%   val IR: EffectExp with FutureOpsExp
%   import IR._
%   override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
%     case FutureMap(a, x, b) =>
%       emitValDef(sym, "new Promise()")
%       stream.println(quote(a) + ".onComplete(function (" + quote(x) + ") {")
%       emitBlock(b)
%       stream.println(quote(sym) + ".complete(" + quote(getBlockResult(b)) + ");")
%       stream.println("});")
%     // ... same for each node
%     case _ => super.emitNode(sym, rhs)
% \end{figure}
% The \code{JSGenFutureOps} code generator inherits from the \code{JSNestedCodegen} code generator that defines the building blocks of the JavaScript code generation. The main method is \code{emitNode}, it is called by LMS for each traversed node of the statement graph. In the listing we only showed the code generation for the \code{FutureMap} node: when the asynchronous value on which the \code{map} method has been applied is available, we apply the function passed to the \code{map} method on the value and complete a newly created asynchronous value with the result. The \code{Promise} constructor is provided by a JavaScript preamble and creates an asynchronous value with two methods: \code{onComplete}, registering a callback to be called when the value is available, and \code{complete}, fulfilling the promise with a value.

\subsection{DOM definition}

In this section we show how we can define, as an embedded DSL, with minimal effort, a template engine that is
statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language,
requires no extra compilation step and that can be used on both client-side and server-side.

@@

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and
generates a code equivalent to the listing \ref{dom-api}:

@@

For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on
server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):

@@

We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
of our DSLs: dynamic content of templates is written using embdedded DSLs too, so their translation into JavaScript
and Scala is managed by their respective code generators.

\subsection{More general purpose improvements of JavaScript}

This section presents some more general purpose improvements of JavaScript.

\subsubsection{Null references}

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider the following typical code finding a particular widget in the page and a then particular button in
the widget:

Our solution has both the safety and performance of the listing \ref{null-defensive} and the expressiveness of the
listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the
\code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead the
generated code just checks if a value is \code{null} or not. Here is a listing that uses our DSL and compiles to code
equivalent to the listing \ref{null-defensive}:

\subsubsection{State monad}

Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation
constructs. For example, consider the following function incrementing a counter and returning its previous value:
\begin{lstlisting}[language=JavaScript,label=state-fragile,caption=Fragile state handling]
var x;
var inc = function () {
  var prev = x;
  x = x + 1;
  return prev
};

// Usage
x = 42;
inc();
var y = inc();
alert(y);
\end{lstlisting}
This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
without first initializing the \code{x} variable the behavior will be unexpected.

To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a
factory function. Thus the state is hidden and developers may not forget the initialization:

\begin{lstlisting}[language=JavaScript,label=state-encapsulated,caption=State encapsulation within a function]
var incFactory = function (init) {
  var x = init;
  return function () {
    var prev = x;
    x = x + 1;
    return prev
  }
};

// Usage
var inc = incFactory(42);
inc();
var x = inc();
alert(x);
\end{lstlisting}

The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in
the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}. Getting rid of
mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and
returns its new value in its result:

\begin{lstlisting}[language=JavaScript,label=state-explicit,caption=Explicit state threading]
var inc = function (state) {
  return {
    state: state + 1,
    value: state
  }
};

// Usage
var x1 = inc(42);
var x2 = inc(x1.state);
alert(x2.value);
\end{lstlisting}

However, explicitly threading the state in all functions parameters is not desired because of the syntactic and
performance cost of adding an extra parameter to each function and of returning two results instead of one.
Furthermore, manually passing the result of each previous call to the next one is error prone.

The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids explicitly threading the state
in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
computations, and thus to implicitly thread a state across computations, as shown in the following listing:

\begin{lstlisting}[label=state-monad,caption=Implicitly threaded state using the State Monad]
val inc = for {
  x <- get[Int]
  _ <- put(x + 1)
} yield x

// Usage
val usage = for {
  _ <- inc
  x <- inc
} yield alert(x)
usage.init(42)
\end{lstlisting}

The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state
and to change it. The \code{usage} value is a reusable piece of program that calls two times the \code{inc} function
and prints the result of the second call. The last line of the listing runs the \code{usage} program on an initial
state.

The State monad programming model has the advantages but not the shortcomings of both listings
\ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly
threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives)
and everything is immutable.

Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the
state through. To achieve that, we transform the intermediate representation of the stack of state modifications
produced by the \code{for} sequences into an equivalent intermediate representation using regular variables. Finally,
the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no
abstraction penalty.

\subsubsection{\emph{Adhoc} polymorphism}

>>>>

\title{Statically Typed Web Programming using Scala}

\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered}\textsuperscript{\textdaggerdbl} \and Olivier
\noun{Barais}\textsuperscript{\textasteriskcentered} \and Tiark \noun{Rompf}\textsuperscript{\textdagger} \and
Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France:
\texttt{\{first\}.\{last\}@irisa.fr}\\
\textsuperscript{\textdaggerdbl}Zenexity, Paris, France: \texttt{jrf@zenexity.com}\\
\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{\{first\}.\{last\}@epfl.ch}}

@@

\begin{abstract}
Rich Internet Applications involve even more code on the client-side, dealing with DOM manipulation, event handling
and asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
language has no static typing and no module system. We investigate further the definition of JavaScript as a Scala
compiled embedded DSL using Lightweight Modular Staging. We define safe and expressive abstractions for DOM
manipulation, event handling and asynchronous programming that translate to efficient JavaScript code using native
APIs. We use both type-level information and staging to introduce high-level abstractions with few or no overhead on
the generated JavaScript code.
\end{abstract}

@@

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
writing distributed code requires its different parts to be consistent and leads to code duplication unless one uses
the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole
application. However this language has several drawbacks making it hardly suitable for large code base (\eg{} no
static typing, no module system, verbose syntax, \etc). Some other in-browser execution environments give the
opportunity to write the client-side code in another language that JavaScript, \eg{} Java
applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft
Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
these technologies also have several drawbacks: they require an additional browser plugin to be installed (which may
not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
structured in URLs that users can bookmark or share.

An increasing number of initiatives attempt to allow developers to write the client-side code in a language different
than JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). These languages are
usually more suitable to write large applications either because they provide more constructs to build abstractions
(such as object orientation), or because they are statically typed, or because they add some useful concepts
that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile
for both server and client sides usually allows developers to share some parts of code between server and client
sides for free. However this shared code is restricted to exclusively use language constructs: concepts provided by
external APIs can't be shared between server and client sides because the bindings with these environments are
different on the server and client sides.

In this paper we investigate further \emph{js-scala}, a path already introduced by Kossakowski \etal{}: defining
JavaScript as a compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}. This approach is based on Lightweight
Modular Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate
representation that can be further processed to perform domain specific optimizations and then to generate a
JavaScript program (for the client-side) and a Scala program (for the server-side). This approach has two main
advantages: (1) the embedding reduces the effort needed to define the language since we can reuse the Scala
infrastructure and tooling, and (2) staging gives more knowledge to the compiler about how a given abstraction
should be efficiently translated into the target environment. In other words any library-level abstraction can have
the efficiency of a language-level abstraction.

This previous work on js-scala showed how to write client-side code in Scala using the JavaScript embedded DSL and
how to share code between server-side and client-side. The authors also showed that the code was safer and more
convenient to write thanks to Scala's static typing and class system. Finally, they showed how to \emph{un-invert}
the control of callback-based asynchronous APIs by using continuations but they did not address other concerns of
RIAs development such as user interface events handling and DOM creation and manipulation.

This paper continues their work and presents safe and expressive DSL constructs for RIAs development. We increase the
expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs. We
make asynchronous programming simpler by raising asynchronous computations to first-class citizen. We leverage the
Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type coercion to
the client-side.

@@

The next section introduces a simple application example showing challenges of RIA development, section
\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution, section
\ref{related} presents related works and compares it to related work and section \ref{discussion} concludes.

@@

\section{Background and Motivating Example}
\label{problem}

To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind
mapping” application and show how the client-side code would typically be written in JavaScript.

\subsection{Example Description}

In our application, a mind map is represented as a graph where ideas or concepts are represented by vertices, and the
relation between two ideas is represented by an edge. Users can visualize a mind map, zoom in and out and move it.
They can also add ideas and link them together. Figure \ref{mindmap} shows the visualization of a simple map
consisting of three ideas linked together, using our application.

@@

\subsubsection{Event handling}

@@

We claim this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
the listened event but developers have no way to check that the fields they’re using are indeed defined on the event
they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the
\code{mousewheel} event.

@@

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or
Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}}
aim to make it simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML
structure and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
client-side and server-side, however none has a practical expression language (dynamic content can only be provided
as a map of key-value pairs and no operation can be done on values) or is statically typed.

@@

\subsubsection{Lack of static hecks}

Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
and robust code bases and to perform refactorings. For instance, in the case of RIAs, a mispell in an event name may
break the program behavior without anyone getting a sensible error message.

@@

Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
hampers modularization. Furthermore, native APIs are too verbose, making the code hard to read.

@@

\section{A DSL for Web Programming}
\label{solution}

@@

\subsubsection{Overview}

@@

\subsubsection{Example: null references}

As an illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL provides an
abstraction at the stage-level that is removed by optimization during the code generation.

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider the following typical JavaScript code finding a particular widget in the page and a then particular
button in the widget:

We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
Here is a Scala listing that uses our DSL (implementation details are given in section \ref{implementation}):

The evaluation of the above listing produces a graph of statements from which JavaScript code equivalent to
listing \ref{null-defensive} is generated.

The remaining of this section describes the design and the implementation of DSLs for Web programming using LMS.

\subsection{Event Handling}

We want to bring safety to the event handling API so that developers cannot mispell an event name and cannot attempt
to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
type of the event passed to the handler varies with the event name. For instance in the listing \ref{event-js}, the
handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
\code{MouseWheelEvent} have a property \code{wheelDeltaY} but that’s not the case of an event data of type
\code{KeyboardEvent}, for example.

@@

However, implementing this solution requires a high effort because we have to define as many methods as there are
events. We want to provide a single polymorphic method similar to the native API. To achieve that, we need to encode
the dependency relation between an event name and its corresponding event type at the type-level. This dependency can
naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:

@@

\subsection{Asynchronous Programming}

We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
turning them into first-class citizens. The DSL is monadic so we can solve the callback-hell problem thanks to the
Scala \code{for} notation.

@@

@@

An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
\begin{itemize}
\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} also eventually transforms the value when available~;
\end{itemize}

Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class
citizens: functions can take them as parameters and return them.

@@

Listing \ref{async-parallel-1} defines two asynchronous computations running in parallel and listing
\ref{async-parallel-2} shows the generated JavaScript code (we renamed some identifiers for the sake of readability).
The generated \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to
call when the asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a
\code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when
received.

\subsection{DOM Definition}

In this section we show how we can define a template engine as an embedded DSL, with minimal effort. This template
engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
language, requires no extra compilation step and can be used on both client-side and server-side.

@@

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and
generates a code equivalent to listing \ref{dom-api}:

@@

For instance, the template written in listing \ref{forest-hello} produces the following Scala code usable on
server-side (the generated code for client-side is roughly equivalent to listing \ref{dom-api}):

@@

We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
and Scala is managed by their respective code generators.

\subsection{Leveraging the Scala Type System on the Client-Side}

This section presents some more general purpose improvements of JavaScript based on type-directed mechanisms of
Scala.

% \subsubsection{State monad}
% Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation
% constructs. For example, consider the following function incrementing a counter and returning its previous value:
% \begin{lstlisting}[language=JavaScript,label=state-fragile,caption=Fragile state handling]
% var x;
% var inc = function () {
%   var prev = x;
%   x = x + 1;
%   return prev
% };
% // Usage
% x = 42;
% inc();
% var y = inc();
% alert(y);
% This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
% without first initializing the \code{x} variable the behavior will be unexpected.
% To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a
% factory function. Thus the state is hidden and developers may not forget the initialization:
% \begin{lstlisting}[language=JavaScript,label=state-encapsulated,caption=State encapsulation within a function]
% var incFactory = function (init) {
%   var x = init;
%   return function () {
%     var prev = x;
%     x = x + 1;
%     return prev
% };
% // Usage
% var inc = incFactory(42);
% inc();
% var x = inc();
% alert(x);
% The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in
% the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}. Getting rid of
% mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and
% returns its new value in its result:
% \begin{lstlisting}[language=JavaScript,label=state-explicit,caption=Explicit state threading]
% var inc = function (state) {
%   return {
%     state: state + 1,
%     value: state
% };
% // Usage
% var x1 = inc(42);
% var x2 = inc(x1.state);
% alert(x2.value);
% However, explicitly threading the state in all functions parameters is not desired because of the syntactic and
% performance cost of adding an extra parameter to each function and of returning two results instead of one.
% Furthermore, manually passing the result of each previous call to the next one is error prone.
% The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids explicitly threading the state
% in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
% computations, and thus to implicitly thread a state across computations, as shown in the following listing:
% \begin{lstlisting}[label=state-monad,caption=Implicitly threaded state using the State Monad]
% val inc = for {
%   x <- get[Int]
%   _ <- put(x + 1)
% } yield x
% // Usage
% val usage = for {
%   _ <- inc
%   x <- inc
% } yield alert(x)
% usage.init(42)
% The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state
% and to change it. The \code{usage} value is a reusable piece of program that calls twice the \code{inc} function
% and prints the result of the second call. The last line of the listing runs the \code{usage} program on an initial
% state.
% The State monad programming model has the advantages but not the shortcomings of both listings
% \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly
% threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives)
% and everything is immutable.
% Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the
% state through. To achieve that, we transform the intermediate representation of the stack of state modifications
% produced by the \code{for} sequences into an equivalent intermediate representation using regular variables. Finally,
% the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no
% abstraction penalty.

\subsubsection{\emph{Ad hoc} polymorphism}
